---
title: "The spray package: sparse arrays in R"
output:
  github_document:
    pandoc_args: --webtex
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```


<!-- badges: start -->
[![Build Status](https://travis-ci.org/RobinHankin/spray.svg?branch=master)](https://travis-ci.org/RobinHankin/spray)
[![CRAN_Status_Badge](https://www.r-pkg.org/badges/version/spray)](https://cran.r-project.org/package=spray)
[![Rdoc](http://www.rdocumentation.org/badges/version/spray)](http://www.rdocumentation.org/packages/spray)
<!-- badges: end -->

# Overview

The ```spray``` package provides functionality for sparse arrays.  
In a sparse arrays, nonzero elements are stored along with an index
vector describing their coordinates.  The `spray` package provides
functionality for sparse arrays and interprets them as multivariate
polynomials.


# Installation

You can install the released version of `spray` from
[CRAN](https://CRAN.R-project.org) with:

```{r, message=FALSE}
# install.packages("spray")  # uncomment this to install the package
library("spray")
```

# The `spray` package in use

Base R has extensive support for multidimensional arrays.  Consider

```{r}
a <- array(0,dim=4:12)
a[2,2,2,2,2,2,2,2,2] <- 17
a[3,4,2,2,7,2,3,2,3] <- 18
```

Handling `a` requires storage of $\prod_{i=4}^{12}i=79833600$ floating
point numbers (of which two are nonzero), represented in an elegant
format amenable to extraction and replacement.  Arrays such as this in
which many of the elements are zero are common and in this case
storing only the nonzero elements and their positions would be a more
compact and efficient representation.  To create a sparse array object
in the `spray` package, one specifies a matrix of indices with each
row corresponding to the position of a nonzero element, and a numeric
vector of values:
	
```{r}
library("spray")
M <- rbind(
  c(2,2,2,2,2,2,2,2,2),
  c(3,4,2,2,7,2,3,2,3))

S1 <- spray(M,7:8)
S1
``` 

Note that object `S1` is rather compact by comparison with plain
array `a`, as it needs to record only a 18-element index array of
integers and two double-precision entries.  The order in which the
elements are stored is implementation-specific (see the vignette for
details and an extended discussion).

Basic arithmetic is implemented where appropriate.  If we define

```{r}
S2 <-spray(rbind(
  c(1,2,3,1,3,3,1,4,1),
  c(3,4,2,2,7,2,3,2,3)), c(100,-8))
S2
```

then 

```{r}
S1+S2
```

(the entry with value `8` has cancelled out).

# The spray package and multivariate polynomials

One natural application for `spray` objects is multivariate
polynomials.  Defining

```{r}
S1 <- spray(matrix(c(0,0,0,1,0,0,1,1,1,2,0,3),ncol=3),1:4)
S2 <- spray(matrix(c(6,-7,8,0,0,2,1,1,3),byrow=TRUE,ncol=3),c(17,11,-4))
S1
S2
```

it is natural to interpret the rows of the index matrix as powers of
different variables of a multivariate polynomial, and the values as
being the coefficients.  This is realised in the package using the
`polyform` print option, which if set to `TRUE`, modifies the print
method:

```{r}
options(polyform = TRUE)
S1
S2
```

(only the print method has changed; the objects themselves are
unaltered).  The print method interprets, by default, the three
columns as variables $x,y,z$ although this behaviour is
user-definable.  With this interpretation, multiplication and addition
have natural definitions as multivariate polynomial multiplication and
addition: 


```{r}
S1+S2
S1*S2
S1^2+4*S2
```

It is possible to introduce an element of symbolic calculation,
exhibiting familiar algebraic identities.  Consider the `lone()`
function, which creates a sparse array whose multivariate polynomial
interpretation is a single variable:

```{r}
x <- lone(1, 3)
y <- lone(2, 3)
z <- lone(3, 3)
(x + y) * (y + z) * (x + z) - (x + y + z) * (x*y + x*z + y*z)
```

thus illustrating the identity
$(x+y+z)(xy+xz+yz)=(x+y)(y+z)(x+z)+xyz$.

Spray objects can be coerced to functions:

```{r}
S4 <- spray(cbind(1:3, 3:1), 1:3)
f <- as.function(S4)
f(c(1, 2))
```

Differentiation is also straightforward.  Suppose we wish to calculate
the multivariate polynomial corresponding to
 
\[
\frac{\partial^6}{\partial x\,\partial^2y\,\partial^3z}
\left(xyz + x+2y+3z\right)^3.
\]

This would be

```{r}
aderiv((xyz(3) + linear(1:3))^3, 1:3)
```

The package vignette offers a detailed discussion of the package
design philosophy; also, the `mvp` package provides a further
interpretation of the concept of "sparse" in the context of
multivariate polynomials.
